Supplementary Information

### 2021-02-09 doc/manuscript/MEE_manuscript.Rmd
finish Appendix 1 plot descriptions

[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Ross Alexander^5^,
Craig D. Allen^6^,
Raquel Alfaro-Sánchez^7^,
Tala Awada^8^,
Jennifer L. Baltzer^7^,
Patrick J. Baker^9^,
Sarayudh Bunyavejchewin^10^,
Paolo Cherubini^11^,
Justin Cooper^12^,
Stuart J. Davies^2^,
Cameron Dow^1,13^,
Ryan Helcoski^1^,
[Jakub Kašpar](0000-0003-1780-6310)^14^,
[James Lutz]( 0000-0002-2560-0710)^12^,
Ellis Q. Margolis^6^,
Justin Maxwell^15^,
Sean McMahon^2,16^,
Camille Piponiot^1,2, 17^,
Sabrina Russo^8^,
[Pavel Šamonil](0000-0002-7722-8797)^14^,
Anastasia Sniderhan^7^,
Alan J. Tepley^1,18^,
[Ivana Vašíčková](0000-0002-6070-5956)^14^, 
Mart Vlam^19^,
Peter Zuidema^19^

@@

12. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA

@@

15. Department of Geography, Indiana University, Bloomington, Indiana, USA

16. Smithsonian Environmental Research Center, Edgewater, MD, USA

17. CIRAD, Montpellier, France

18. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

19. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands

>>>>

[Kristina J. Anderson-Teixeira^1,2^*](0000-0001-8461-9713),
Valentine Herrmann^1^,
Christy Rollinson^3^,
Bianca Gonzalez^1^,
Erika B. Gonzalez-Akre^1^,
Neil Pederson^4^,
Ross Alexander^5^,
Craig D. Allen^6^,
Raquel Alfaro-Sánchez^7^,
Tala Awada^8^,
Jennifer L. Baltzer^7^,
Patrick J. Baker^9^,
Sarayudh Bunyavejchewin^10^,
Paolo Cherubini^11^,
Justin Cooper^12^,
Stuart J. Davies^2^,
Cameron Dow^1,13^,
Ryan Helcoski^1^,
[Jakub Kašpar](0000-0003-1780-6310)^14^,
[James Lutz]( 0000-0002-2560-0710)^15^,
Ellis Q. Margolis^6^,
Justin Maxwell^16^,
Sean McMahon^2,17^,
Camille Piponiot^1,2, 18^,
Sabrina Russo^8^,
[Pavel Šamonil](0000-0002-7722-8797)^14^,
Anastasia Sniderhan^7^,
Alan J. Tepley^1,19^,
[Ivana Vašíčková](0000-0002-6070-5956)^14^, 
Mart Vlam^20^,
Peter Zuidema^20^

@@

12. University of Alberta; Edmonton, Alberta, Canada

@@

15. S. J. & Jessie E. Quinney College of Natural Resources and the Ecology Center, Utah State University, Logan, UT 84322, USA

16. Department of Geography, Indiana University, Bloomington, Indiana, USA

17. Smithsonian Environmental Research Center, Edgewater, MD, USA

18. CIRAD, Montpellier, France

19. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada

20. Forest Ecology and Forest Management Group, Wageningen; Wageningen, The Netherlands

### 2021-02-09 doc/manuscript/MEE_manuscript_SI.Rmd
finish Appendix 1 plot descriptions

*(include descriptions of stand history, global change dynamics)*

@@

### Cedar Breaks, Utah, USA

has tree-ring records dating back to 0 AD (Cooper & Lutz, unpublished data)

### Scotty Creek, Northwest Territories, Canada
melting permafrost

>>>>

@@

### Cedar Breaks, Utah, USA
Cores were collected within the Utah Forest Dynamics Plot ForestGEO plot on the Colorado Plateau in Cedar Breaks National Monument [CB; @furniss_reconciling_2017]. 
The plot sits on the edge of cliff, which descends ~750 m to a valley below.
The mature spruce-fir forest is dominated by the species analyzed here (*Abies bifolia*, *Populus tremuloides*, *Picea engelmannii*, *Pinus flexilis*, *Pinus longaeva*, *Picea pungens*, and *Pseudotsuga menziesii*).
The site is largely undisturbed. 
The most recent fire occurred in 1802, and many currently living trees established before then. Living *Pinus longaeva* at the site date back ~1600 years, and deadwood at the site dates back to 0 AD (Cooper & Lutz, unpublished data).

In 2019, living individuals of each dominant species were cored.  
Chronologies are published on ITRDB [@birch_birch_2020; @birch_birch_2020-2; @birch_birch_2020-3; @birch_birch_2020-4].

### Scotty Creek, Northwest Territories, Canada
The Scotty Creek site is a black spruce (*Picea mariana*) forest on rapidly thawing discontinuous permafrost [@sniderhan_growth_2016; @dearborn_permafrost_2020].
Vertical permafrost thaw has been causing drought stress in the shallow-rooted trees, resulting in growth declines [@sniderhan_growth_2016] and high mortality [@dearborn_permafrost_2020].

In the summers of 2012 and 2013, seven forest stands within a 2 km^2^ area were sampled.
In five stands with clearly defined boundaries at edges of permafrost plateaus, all living and dead trees $\ge$ 7 cm DBH were cored. 
In two stands without well-defined edges, cores were taken from a >7 cm DBH tree nearest to every 20 m interval in a 100 × 100 m grid, totaling 36 trees per stand.
Sampling details are given in @sniderhan_growth_2016.
